---
title: "Reproducible Research - Week 2 - Course Project"
author: "Sourav Datta"
date: "6/19/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Getting and cleaning data

Download the data from the given URL and verify it is in proper format to continue analysis.

```{r}
if (!file.exists('./data')) {
  dir.create('data')
}

download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip', 
              method = 'curl',
              mode = 'wb',
              destfile = 'data/activitydata.zip')

unzip('data/activitydata.zip', overwrite = TRUE, exdir = 'data')

stepdata <- read.csv('data/activity.csv', header = TRUE)
str(stepdata)
```

### 2. Total number of steps taken each day

```{r}
library(dplyr)

stepsbydate <- stepdata %>%
  select(steps, date) %>%
  na.omit() %>%
  group_by(date) %>%
  summarise(nsteps = sum(steps))

hist(stepsbydate$nsteps, xlab = 'Total daily steps', 
     main = 'Histogram of total number of steps by date',
     breaks = 20)
```

### 3. Mean and median of the total number steps taken per day

```{r}
mean1 <- mean(stepsbydate$nsteps)
```

```{r}
median1 <- median(stepsbydate$nsteps)
```


### 4. Time series plot of the average number of steps taken

```{r}
library(ggplot2)

stepsbyinterval <- stepdata %>%
  select(steps, interval) %>%
  na.omit() %>%
  group_by(interval) %>%
  summarise(nsteps = mean(steps))

ggplot(stepsbyinterval, aes(x = interval, y = nsteps)) + geom_line()
```

### 5. The 5-minute interval that, on average, contains the maximum number of steps

```{r}
stepsbyinterval[which(stepsbyinterval$nsteps == max(stepsbyinterval$nsteps)), ]
```

### 6. Imputing missing data

Missing values (i.e. `NA`) can be replaced by the mean of that interval.

```{r}
replace.missing <- function (x) {
  replace(x, is.na(x), mean(x, na.rm = TRUE))
}
```

We can now replace the missing values for each group of interval by the above function.

```{r}
replaced.stepdata <- stepdata %>%
  group_by(interval) %>%
  mutate(steps = replace.missing(steps)) %>%
  ungroup()

str(replaced.stepdata)
```

### 7. Histogram of the total number of steps taken each day after missing values are imputed

Now we recalculate the number of steps each day with replaced data.

```{r}
stepsbydate2 <- replaced.stepdata %>%
  select(steps, date) %>%
  group_by(date) %>%
  summarise(nsteps = sum(steps))

head(stepsbydate2)
```

Summary of new dataset
```{r}
summary(stepsbydate2)
```

Draw histogram with new data

```{r}
hist(stepsbydate2$nsteps, 
     xlab = 'Total daily steps', 
     main = 'Histogram of total number of steps by date',
     breaks = 20)
```

Also, recalculate mean and median of the total number steps taken per day.

```{r}
mean2 <- mean(stepsbydate2$nsteps)
```

```{r}
median2 <- median(stepsbydate2$nsteps)
```

Do, the old and new values of mean and median differ?

```{r}
data.frame(mean_diff = mean2 - mean1, median_diff = median2 - median1)
```

### 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

First, we add a new column which indicates the day is `weekend` or `weekday`.

```{r}
stepsdatawithday <- replaced.stepdata %>%
  mutate(day = weekdays(as.Date(date))) %>%
  mutate(day = ifelse(day == 'Saturday' | day == 'Sunday', 'weekend', 'weekday'))

table(stepsdatawithday$day)
```

Then, we calculate the average number of steps grouped by interval

```{r}
plotdata <- stepsdatawithday %>%
  select(steps, interval, day) %>%
  group_by(day, interval) %>%
  summarise(nsteps = mean(steps))

ggplot(plotdata, aes(x = interval, y = nsteps, color = day)) + geom_line() +
  facet_grid(day ~ .) + xlab('Interval') + ylab('Average daily steps') + 
  ggtitle('Average daily steps per interval')
```

